Towards a Complete Model of Mutation Selection Balance in Humans

### 2023-05-17 manuscript/manuscript.tex
more theory

However, unlike non-heritable variation in offspring number, the heritable
variation is autocorrelated across generations. At the individual level
Robertson considered, this is because offspring from large families tend to
beget many descendents themselves (and likewise with small families). However,
the same process acts at the genomic level, as the perturbations to a neutral
allele's trajectory from its particular fitness background tend to occur in the
same direction across generations. The additional heritable variation in
offspring number is the additive fitness variation $V_A$, which is inflated by
a factor $Q_t^2$ due to the build up of autocorrelation to generation $t$. We
let $Q^2 = Q_t^2$ as $t \to \infty$. Intuitively, the product $V_A Q^2$
represents the expected total variance in reproductive success a neutral
mutation experiences over its lifetime in a system with draft. Incorporating
the impact of time-invariant autocorrelation created by draft as rescaled rate
of drift is analogous to using the Green--Kubo relation to find the transport
coefficient \parencite{Green1954-kl,Kubo1957-va}.

@@

(c.f. \cite{Robertson1961-ho,Santiago1995-hx}; see XXX for a proof). In
general, however, statistics such as heterozygosity depend on the cumulative
autocorrelation factor \emph{up to some time} $t$, $Q_t^2$ \parencite[p.
2111]{Santiago1998-bs}. This reflects the fact that effective population size
experienced by a mutation newly associated with a fitness background is larger
than the asymptotic effective population size, since not much autocorrelation
has accumulated in the stochastic perturbations caused by drift yet. From a
backwards-time perspective, this reflects the fact that pairwise diversity
(which is approximately heterozygosity when $2N\mu \ll 1$) depends on the
pairwise coalescence rate per generation, which is not constant under draft.
The benefit of using Robertson's forward-time model of draft is that the
inflation factor is invariant with respect to the particular fitness background
the neutral allele becomes stochastically associated with. By contrast, the
difficulty with modeling draft backwards in time is that the coalescence rates
experienced by a lineage are not invariant to which lineage was sampled due to
its particular associated fitness background.

@@

where the factor of one-half comes from ignoring weak associations from
unlinked regions and chromosomes (see XXX). In our genome-wide model, we
consider the summation in Equation \eqref{eq:polygenic_Ne} over non-overlapping
segments $i \in \{1, 2, \ldots, n\}$ each undergoing selection such that
segment $i$ contributes additive fitness variance $V_{A,i}$ to the total
additive genetic fitness variance. Under draft, the autocorrelation function
for a neutral allele associated with segment $i$ is the time-invariant function
$C(t) = \[(1-r_i)(1-\kappa_i)\]^t$, where $r_i$ is the recombination fraction
to the segment and $\kappa_i$ is the rate that the associated fitness decays
due to selective dynamics. Then, the cumulative autocorrelation is,

@@

(see Appendix Equation \ref{eq:Qinf}). This general equation can accommodate
models of polygenic selection as long as the equilibrium additive fitness
variation $C_i^2$ can be specified and the change in variance due to selection
can be approximated as a geometric decay, i.e. $\Delta C_i^2 = -\kappa C_i^2$
\parencite{Bulmer1971-ae,Keightley1988-eq,Walsh2018-bt}. This is usually a
reasonable assumption since within-generation selection removes a fraction of
phenotypic variation from the population, and some fraction of that is additive
genetic variation \parencite{Bulmer1971-ae,Keightley1988-eq}.

The final pieces are the equilibrium additive fitness variance $V_A$ and the
decay rate in associated fitness $1-\kappa$. At equilibrium,

@@

\begin{align}
  \var(x_2 - p_0) &= \E({S_1}^2) + \E(S_1 S_2) + \E({S_2}^2) \\
                  & + \E({D_1}^2) + \E({H_1}^2) + \E({D_2}^2) + \E({H_2}^2) \\
  \frac{\var(x_2 - p_0)}{p_0(1-p_0)} &= \frac{V_h}{2N_{e,1}} + \frac{V_h Q_2^2}{2N_{e,2}}\left(1-\frac{1}{2N_{e,1}}\right)  +  \frac{V_h}{2N_{e,2}}  \left(1-\frac{1}{2N_{e,1}}\right)  \\ 
                                     &+ \frac{V_n}{8N_{e,1}} + \frac{1}{4N_{e,1}} + \frac{V_n}{8N_{e,2}}  \left(1 - \frac{1}{2N_{e,1}}\right) 
+ \frac{1}{4N_{e,2}} \left(1 - \frac{1}{2N_{e,1}}\right).
\end{align}

@@

\begin{align}
  V_{A} = Us - 2Rs. 
\end{align}

where $U = 2u$ and $R$ is the diploid rate of the ratchet, which depends on
higher moments of the fitness distribution, but only differs from $r$ through a
rescaling of the population size. Note that if selected sites are sufficiently
deleterious that they cannot fix (or the population is infinite), $R=0$ and the
additive fitness variance is identical to that under strong background
selection, $V_A = Us = V_\text{BGS}$ (equation \eqref{eq:va_bgs}). Then, the
fixation probability of a new deleterious mutation with additive effects ($1-s$
in the heterozygote, $1-2s$ in the homozygote) in a diploid population is,

>>>>

However, unlike non-heritable variation in offspring number, the heritable
variation is autocorrelated across generations. At the individual level
Robertson considered, this is because offspring from large families tend to
beget many descendents themselves (and likewise with small families). The same
autocorrelation occurs at the genomic level, as the perturbations to a neutral
allele's trajectory from its particular fitness background tend to occur in the
same direction across generations as long as the linkage exists and the
selective environment is constant. The additional heritable variation in
offspring number is the additive fitness variation $V_A$, which is inflated by
a factor $Q_t^2$ due to the build up of autocorrelation to generation $t$. We
let $Q^2 = Q_t^2$ as $t \to \infty$. Intuitively, the product $V_A Q^2$
represents the expected total variance in reproductive success a neutral
mutation experiences over its lifetime in a system with draft. Incorporating
the impact of time-invariant autocorrelation created by draft as a rescaled
rate of drift is analogous to using the Green--Kubo relation to find the
transport coefficient in molecular dynamics with velocity autocorrelation
\parencite{Green1954-kl,Kubo1957-va}.

@@

(c.f. \cite{Robertson1961-ho,Santiago1995-hx}; see XXX for a proof). In
general, however, statistics such as heterozygosity depend on the cumulative
autocorrelation factor \emph{up to some time} $t$, $Q_t^2$ \parencite[p.
2111]{Santiago1998-bs}. This reflects the fact that effective population size
experienced by a mutation newly associated with a fitness background is larger
than the asymptotic effective population size, since not much autocorrelation
has accumulated in the stochastic perturbations caused by drift yet. From a
backwards-time perspective, this reflects the fact that pairwise diversity
(which is approximately heterozygosity when $2N\mu \ll 1$) depends on the
pairwise coalescence rate per generation, which is not constant under draft.

@@

where the factor of one-half comes from ignoring weak associations from
unlinked regions and chromosomes (see XXX). In our genome-wide model, we
consider the summation in Equation \eqref{eq:polygenic_Ne} over non-overlapping
segments $i \in \{1, 2, \ldots, n\}$ each undergoing selection such that
segment $i$ contributes additive fitness variance $V_{A,i}$ to the total
additive genetic fitness variance. Under equilibrium levels of fitness
variation, the autocorrelation function for a neutral allele associated with
segment $i$ is the time-invariant function $C(t) = \[(1-r_i)(1-\kappa_i)\]^t$,
where $r_i$ is the recombination fraction to the segment and $\kappa_i$ is the
rate that the associated fitness variance decays due to selective dynamics.
Then, the cumulative autocorrelation is,

@@

(see Appendix Equation \ref{eq:Qinf}). This general equation can accommodate
models of polygenic selection as long as the equilibrium additive fitness
variation $V_{A,i}$ can be specified and the change in variance due to
selection can be approximated as a geometric decay, i.e. $\Delta V_{A,i} =
-\kappa V_{A,i}$ \parencite{Bulmer1971-ae,Keightley1988-eq,Walsh2018-bt}. This
is usually a reasonable assumption since within-generation selection removes a
fraction of phenotypic variation from the population, and some fraction of that
is additive genetic variation \parencite{Bulmer1971-ae,Keightley1988-eq}.

The remaining pieces are expressions for the equilibrium additive fitness
variance $V_A$ and the decay rate in associated fitness $1-\kappa$. At this
point, we diverge from Santiago and Caballero (\citeyear{Santiago1998-bs},
\citeyear{Santiago2016-mu}) to note that the additive fitness variation is the
sum of additive \emph{genic} fitness variance $V_a$ (i.e. due to allele
frequencies) and the total linkage disequilibria $\delta_{LD}$ between selected
sites, $V_A = V_a + \delta_{LD}$. Considering only the genic variation, at
equilibrium $\Delta V_{a,i} = 0$; thus, the loss in genic fitness each
generation due to selective dynamics $-\kappa V_{A,i}$ must be equal to the
increase in fitness variation to new mutations $V_m$ each generation and
changes in the LD term \parencite{Bulmer1971-ae}. Note that the loss in genic
fitness due to drift is not time-invariant due to the build up of
autocorrelation, and considered elsewhere in the derivation (see Supplementary
Materials Equation \ref{eq:vardecay}). Similarly, the loss in variation due to
fixation of weakly deleterious occurs on much slower timescales and can be
ignored here. Then, $\kappa_i = \nicefrac{V_{m,i}}{V_{a,i}}$ (see Supplementary
Materials Equation \ref{eq:Z}).

Under any selection model, the genic fitness variance created by a new mutation
(at frequency $x=\nicefrac{1}{2N}$) is $2s^2 x(1-x) \approx \nicefrac{s^2}{N}$.
For the entire population of $2N$ chromosomes, the fresh variance from mutation
each generation in segment $i$ is $V_{m,i} \approx U_is^2$ where $U_i = 2\mu
L_i$ is the diploid mutation rate per generation within the segment. Under
mutation-selection balance assumed by BGS, an $L_i$-basepair segment has genic
variance $V_{a,i}^{BGS} \approx U_i s$ (see Supplementary Materials Equation
\ref{eq:va_bgs}) and thus $\kappa_i^{BGS} = s$. Combining this with $\kappa_i =
\nicefrac{V_{m,i}}{V_{a,i}} = s$ in Equation \eqref{eq:simple_Ne}, we have

\begin{align}
    N_e = N \exp \left( - \sum_i^n \frac{\mu L_i}{s(1 + r_i(1-s)/s)^2} \right) 
\end{align}

which is exactly the genome-wide model of background selection used in previous
studies \parencite{McVicker2009-ax,Elyashiv2016-vt,Murphy2022-sj} (note that
\cite{Santiago1998-bs} have shown a similar correspondence between their work
and BGS models with recombination from \cite{Nordborg1996-nq} and
\cite{Hudson1995-xc}).

When selection is weak, however, the additive fitness variance under the
classic BGS model is no longer accurate. \parencite{Santiago2016-mu} suggested
that this is due to the loss of fitness variation that occurs when a
segregating site fixes and its heterozygosity becomes zero. If we let $R$
represent twice the rate of the ratchet of deleterious mutations in fixations
per generation, each fixation removes exactly the variation mutation puts in
$V_m = Us$, thus under a general equilibrium negative selection process,

\begin{align}
  V_{a} = (U - R)s. 
\end{align}

Our results differ from \parencite{Santiago2016-mu}, since we find from
simulations their expression models the additive \emph{genic} fitness
variation, not additive genetic fitness variation. Intuitively, the equilibrium
genic fitness variation is a balance between the rate of gain of genic fitness
variation due to mutation and its loss due to fixation. Note that since
mutation is the ultimate source of all mutations that fix or are lost, $U \ge
R$.

The benefit of using Robertson's forward-time model of draft is that the
inflation factor is invariant with respect to the particular fitness background
the neutral allele becomes stochastically associated with. By contrast, the
difficulty with modeling draft backwards in time is that the coalescence rates
experienced by a lineage are not invariant to which lineage was sampled due to
its particular associated fitness background.

@@

\begin{align}
    \label{eq:vardecay}
  \var(x_2 - p_0) &= \E({S_1}^2) + \E(S_1 S_2) + \E({S_2}^2) \\
                  & + \E({D_1}^2) + \E({H_1}^2) + \E({D_2}^2) + \E({H_2}^2) \\
  \frac{\var(x_2 - p_0)}{p_0(1-p_0)} &= \frac{V_h}{2N_{e,1}} + \frac{V_h Q_2^2}{2N_{e,2}}\left(1-\frac{1}{2N_{e,1}}\right)  +  \frac{V_h}{2N_{e,2}}  \left(1-\frac{1}{2N_{e,1}}\right)  \\ 
                                     &+ \frac{V_n}{8N_{e,1}} + \frac{1}{4N_{e,1}} + \frac{V_n}{8N_{e,2}}  \left(1 - \frac{1}{2N_{e,1}}\right) 
+ \frac{1}{4N_{e,2}} \left(1 - \frac{1}{2N_{e,1}}\right).
\end{align}

@@

\begin{align}
  V_{A} = Us - Rs. 
\end{align}

where $U = 2u$ and $R=2r$ is the diploid rate of the ratchet, which depends on
higher moments of the fitness distribution, but only differs from $r$ through a
rescaling of the population size. Note that if selected sites are sufficiently
deleterious that they cannot fix (or the population is infinite), $R=0$ and the
additive fitness variance is identical to that under strong background
selection, $V_A = Us = V_\text{BGS}$ (equation \eqref{eq:va_bgs}). Then, the
fixation probability of a new deleterious mutation with additive effects ($1-s$
in the heterozygote, $1-2s$ in the homozygote) in a diploid population is,
